\chapter*{Citation and Figure TODOs}
\addcontentsline{toc}{chapter}{Citation and Figure TODOs}

\section*{Verified bibliography entries already added}

The bibliography includes verified entries for the main E-clustered baseline, the finite-size mesoscopic population theory paper, the Rostami E/I clustered attractor paper, balanced-network references, the original SSN paper, the stochastic SSN variability paper, and the broad stimulus-onset variability-quenching paper.

\section*{Citation TODOs that remain}

\begin{enumerate}[leftmargin=*]
  \item Verify any project-specific Tilo notes, especially the source referred to as \texttt{ssn\_spiking.pdf}. It is intentionally not cited because the file was not available in this coordinator context.
  \item Verify the Huang/Sit/Schwalger/Abbott/Miller/Doiron manuscript status, title, and claims before replacing the conductance-shunting TODOs in Chapter 17.
  \item Check paper-specific parameter tables before turning provisional reproduction ranges in Chapter 21 into cited numerical facts.
  \item Check the exact Rostami et al.\ Methods details before replacing all architecture-specific TODOs with claims about which connection blocks are clustered and how parameter robustness was quantified.
\end{enumerate}

\section*{Figure status}

All eight requested raster figures exist under \texttt{figures/} and are used in the course only as PNG assets. No missing figure slots remain in the integrated \LaTeX{} source. Exact equations and derivations remain in \LaTeX{} rather than embedded in images.
